---
title: 'Supplementary material and reproducible research files for article "*A Joint Bayesian Framework for Measurement Error and Missing Data using Integrated Nested Laplace Approximations*"'
---

Authors: Skarstein, E., Martino, S., Muff, S.

This supplementary material was written by Skarstein, E., but code for analysis was developed jointly between authors.

In case of any questions or comments, please do not hesitate to contact <emma.s.skarstein@ntnu.no>!

This website contains scripts, tutorials and data to reproduce all analysis and figures from the manuscript.

## Examples 
These include scripts to run the code for analysis, including comments and explanations.

- Simulation example
- Simulation study
- Missing value imputation using INLA
- Missing and mismeasured covariates
- Examples in ‘inlabru’

## Datasets
All the data sets that are used in the paper can be downloaded manually from [link to github repo], or like this:

```{r}
# Code to download data
```

## Session info

The code was written and run in R with the following software versions:

```{r}
inla.version()
```


```{r, , echo=FALSE}
#| results: asis
#| echo: false
sessioninfo::session_info(pkgs = c("INLA", "inlabru")) |> 
  details::details(
    summary = 'Current session info',
    open    = TRUE
  )
```

