---
title: "Missing covariate imputation"
execute: 
  freeze: auto
---


In this example, we use the `nhanes2` data set from the `mice` R-package to illustrate how to do missing covariate imputation in INLA by using a measurement error model.

## Loading packages

```{r}
#| label: setup
#| message: false
library(mice)    # Just used for the nhanes2 data set
library(INLA)    # INLA modelling
library(dplyr)   # Data wrangling of the results
library(gt)      # Tables

```

## Loading and preparing the data

```{r}
# Using the nhanes data set found in mice so we can compare to mice:
data(nhanes2)

head(nhanes2)

n <- nrow(nhanes2)

# Manually dummy-code age:
age2 <- ifelse(nhanes2$age == "40-59", 1, 0)
age3 <- ifelse(nhanes2$age == "60-99", 1, 0)

# Center the response and continuous covariates
chl <- scale(nhanes2$chl, scale = FALSE)[,1]
bmi <- scale(nhanes2$bmi, scale = FALSE)[,1]
```

## Modelling

We want to fit the model

$$
chl \sim \beta_0 + \beta_1 age_2 + \beta_2 age_3 + \beta_3 bmi
$$
TODO: Coefficient names are not consistent

### Specifying priors

```{r}
# Priors for model of interest coefficients
prior.beta = c(0, 1/1000) # N(0, 10^3)

# Priors for exposure model coefficients
prior.alpha <- c(0, 1/10000) # N(0, 10^4) 

# Priors for y, measurement error and true x-value precision
prior.prec.y <- c(0.5, 0.5) # Gamma(0.5, 0.5)
prior.prec.u_c <- c(0.5, 0.5) # Gamma(0.5, 0.5)
prior.prec.x <- c(0.5, 0.5) # Gamma(0.5, 0.5) 

# Initial values
prec.y <- 1
prec.u_c <- 1
prec.x <- 1
```


### Setting up the matrices for the joint model

```{r}
Y <- matrix(NA, 3*n, 3)

Y[1:n, 1] <- chl             # Regression model of interest response
Y[n+(1:n), 2] <- bmi         # Error model response
Y[2*n+(1:n), 3] <- rep(0, n) # Exposure model response

beta.0 <- c(rep(1, n), rep(NA, n), rep(NA, n))
beta.x <- c(1:n, rep(NA, n), rep(NA, n))
beta.z2 <- c(age2, rep(NA, n), rep(NA, n))
beta.z3 <- c(age3, rep(NA, n), rep(NA, n))

id.x <- c(rep(NA, n), 1:n, 1:n) 
weight.x <- c(rep(1, n), rep(1, n), rep(-1, n))

alpha.0 <- c(rep(NA, n), rep(NA, n), rep(1, n))
alpha.z2 <- c(rep(NA, n), rep(NA, n), age2)
alpha.z3 <- c(rep(NA, n), rep(NA, n), age3)

dd <- data.frame(Y = Y, 
                 beta.0 = beta.0,
                 beta.x = beta.x,
                 beta.z2 = beta.z2,
                 beta.z3 = beta.z3,
                 id.x = id.x,
                 weight.x = weight.x,
                 alpha.0 = alpha.0,
                 alpha.z2 = alpha.z2,
                 alpha.z3 = alpha.z3)
```

### INLA formula
```{r}
formula = Y ~ - 1 + beta.0 + beta.z2 + beta.z3 + 
  f(beta.x, copy="id.x", 
    hyper = list(beta = list(param = prior.beta, fixed=FALSE))) +
  f(id.x, weight.x, model="iid", values = 1:n, 
    hyper = list(prec = list(initial = -15, fixed=TRUE))) +
  alpha.0 + alpha.z2 + alpha.z3
```

### Scaling of ME precision

Since we are not assuming any measurement error here, we need to "turn off" the error model for the observations where the observation is not missing. 
```{r}
# Scale the measurement error precision according to whether observations are missing or not (1 if missing, large if not missing)
M <- ifelse(is.na(bmi), 1, 10^12)
# Scaling the precision to be large for the error model (since we only have correctly observed or missing data, we want high precision for the correctly observed data and for the missing data it doesn't matter)
Scale <- c(rep(1, n), M, rep(1, n))
```

### Fitting the model
```{r}
model_missing <- inla(formula, data = dd, scale = Scale,
                     family = c("gaussian", "gaussian", "gaussian"),
                     control.family = list(
                       list(hyper = list(prec = list(initial = log(prec.y), 
                                                     param = prior.prec.y, 
                                                     fixed = FALSE))),
                       list(hyper = list(prec = list(initial = log(prec.u_c), 
                                                     param = prior.prec.u_c, 
                                                     fixed = TRUE))),
                       list(hyper = list(prec = list(initial = log(prec.x), 
                                                     param = prior.prec.x, 
                                                     fixed = FALSE)))
                     ),
                     control.fixed = list(
                       mean = list(beta.0 = prior.beta[1], 
                                   beta.z2 = prior.beta[1], 
                                   beta.z3 = prior.beta[1],  
                                   alpha.0 = prior.alpha[1], 
                                   alpha.z2 = prior.alpha[1],
                                   alpha.z3 = prior.alpha[1]), 
                       prec = list(beta.0 = prior.beta[2], 
                                   beta.z2 = prior.beta[2], 
                                   beta.z3 = prior.beta[2],  
                                   alpha.0 = prior.alpha[2], 
                                   alpha.z2 = prior.alpha[2],
                                   alpha.z3 = prior.alpha[2])),
                     verbose=F)
```


```{r}
# Save results:
saveRDS(model_missing, file = "results/model_missing.rds")
```

### Fitting a complete case model

```{r}
# Where is bmi missing? 
missing_bmi <- is.na(bmi)

dd_naive <- data.frame(Y = chl, 
                       beta.0 = rep(1, length(bmi)),
                       beta.x = bmi, 
                       beta.z2 = age2, 
                       beta.z3 = age3)[!missing_bmi, ]


# Priors for model of interest coefficients
prior.beta <- c(0, 1/1000) # N(0, 10^3)

# Formula
formula <- Y ~ - 1 + beta.0 + beta.z2 + beta.z3 + beta.x

# Fit model
model_naive <- inla(formula,
              data = dd_naive,
              family = c("gaussian"),
              control.family = list(
                list(hyper = list(prec = list(initial = log(1), 
                                              param = c(0.5, 0.5), # Gamma(0.5, 0.5) 
                                              fixed = FALSE)))),
              control.fixed = list(
                       mean = list(beta.0 = prior.beta[1], 
                                   beta.z2 = prior.beta[1], 
                                   beta.z3 = prior.beta[1],
                                   beta.x = prior.beta[1]), 
                       prec = list(beta.0 = prior.beta[2], 
                                   beta.z2 = prior.beta[2], 
                                   beta.z3 = prior.beta[2],  
                                   beta.x = prior.beta[2])),
)
```



## Results

The posterior means and standard deviations are presented in the table below.

```{r}
#| echo: false

return_estimates <- function(inla.object){
  fixed <- data.frame(inla.object$summary.fixed)
  fixed <- fixed[, names(fixed) != "kld"]
  fixed$coefficient.name <- rownames(fixed)
  hyperpar <- data.frame(inla.object$summary.hyperpar)
  hyperpar$coefficient.name <- rownames(hyperpar)

  # Extract all coefficients that contain "beta" (the coefficients of the model of interest)
  betas <- rbind(dplyr::filter(fixed, grepl("beta", coefficient.name)),
                 dplyr::filter(hyperpar, grepl("beta", coefficient.name)))
  # Extract all coefficients that contain "alpha" (the coefficients of the imputation model)
  alphas <- dplyr::filter(fixed, grepl("alpha", coefficient.name))
  
  # All coefficients
  all_coefs <- dplyr::bind_rows("Imputation model" = alphas, 
                                "Model of interest" = betas, 
                                .id = "model")
  
  return(all_coefs)
}
```

```{r}
#| echo: false
#| eval: false

return_estimates(model_missing) %>% 
  select(coefficient.name, mean, sd, model) %>% 
  arrange(desc(model)) %>% # Arrange so the MOI coefs are at the top
  group_by(model) %>% # Group by model type
  gt(rowname_col = "coefficient.name")
```


```{r}
#| echo: false
inla_mcmc_results <- tibble::tribble(
  ~coefficient.name, ~model,              ~inla_mcmc.mean, ~inla_mcmc.sd,
  "beta.0",          "Model of interest", 43.469, 62.603,
  "Beta for beta.x", "Model of interest", 4.864, 2.206, 
  "beta.z2",         "Model of interest", 29.501, 17.871, 
  "beta.z3",         "Model of interest", 49.449, 23.207
)

naive_results <- return_estimates(model_naive) |> 
  select(coefficient.name, mean, sd, model) %>% 
  rename(naive.mean = mean, naive.sd = sd)
naive_results$coefficient.name[4] <- "Beta for beta.x"

return_estimates(model_missing) %>% 
  select(coefficient.name, mean, sd, model) %>% 
  rename(me_adjusted.mean = mean, me_adjusted.sd = sd) %>%
  full_join(naive_results, by = c("coefficient.name", "model")) %>% 
  full_join(inla_mcmc_results, by = c("coefficient.name", "model")) %>% 
  mutate_if(is.numeric, round, digits = 3) %>% 
  arrange(desc(model)) %>% # Arrange so the MOI coefs are at the top
  group_by(model) %>% # Group by model type
  gt(rowname_col = "coefficient.name") %>% 
  tab_spanner_delim(
    delim = "."
  )

```


https://stefvanbuuren.name/RECAPworkshop/Practicals/RECAP_Practical_II.html 

